---
title: "KEGGgraph: Application Examples"
author:
  - name: "Jitao David Zhang"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  BiocStyle::html_document
bibliography: KEGGgraph.bib
csl: bioinformatics.csl
reference-section-title: References
link-citations: yes
abstract: >
  In this vignette, we demonstrate the application of `r Biocpkg("KEGGgraph")` as
  flexible module in analysis pipelines targeting heterogenous biological
  questions. For basic use of the `r Biocpkg("KEGGgraph")` package, please refer
  to the vignette *KEGGgraph: a graph approach to KEGG PATHWAY in R and
  Bioconductor*.
vignette: >
  %\VignetteIndexEntry{ KEGGgraph: Application Examples }
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
## The microarray section needs the annotation packages to map Affymetrix probe
## identifiers onto Entrez gene identifiers. They are all suggested rather than
## required, so the section is skipped where they are unavailable and the
## vignette still builds.
microarrayPkgs <- c("hgu133plus2.db", "org.Hs.eg.db", "RBGL", "RColorBrewer")
hasMicroarrayDeps <- all(vapply(microarrayPkgs, requireNamespace,
                                logical(1), quietly = TRUE))

## Lazily evaluated, so that the figures computed above are never forced when
## the annotation packages are missing.
val <- function(expr, fallback = "(not computed)") {
  if (hasMicroarrayDeps) expr else fallback
}
```

# Introduction

In many cases, `r Biocpkg("KEGGgraph")` is used as stand-alone software package
to parse KEGG pathway [@Kanehisa08] into graph models and consequently to
analyze them. However, `r Biocpkg("KEGGgraph")` can also be combined with other
tools (within or out of the scope of **R** and **Bioconductor** [@Gentleman04],
[@Carey05]), to build flexible analysis pipelines. In this vignette we
demonstrate the cooperation between `r Biocpkg("KEGGgraph")` and other tools
with several examples.

# Parse created or edited pathway

KGML-ED [@Klukas07] is a tool designed for the dynamic visualization,
interactive navigation and editing of KEGG pathway diagrams. The program is able
to modify or even create *de novo* new pathways and export them into KGML (KEGG
XML) files. Since `r Biocpkg("KEGGgraph")` captures the graph information of
KEGG PATHWAY by parsing KGML files, it is also capable to parse the KGML files
exported from KGML-ED program. The joint use of the two programs allows user
edit KGML files from KEGG (for example, by adding/removing new nodes/edges), or
create new pathways from scratch, and then analyse them.

Here we demontrate the parse of a toy pathway (network motif) created by
KGML-ED, which illustrates a feed-forward loop (FFL) [@Mangan2003], [@Alon2007].

```{r loadtoy, echo=TRUE, message=FALSE}
library(KEGGgraph)
toyKGML <- system.file("extdata/kgml-ed-toy.xml", package="KEGGgraph")
toyGraph <- parseKGML2Graph(toyKGML, genesOnly=FALSE)
toyGraph
nodes(toyGraph)
```

We visualize the graph with `r Biocpkg("Rgraphviz")` in figure [1](#fig:01).

```{r vistoy, echo=FALSE, results='hide', fig.show='hide', message=FALSE, warning=FALSE}
library(Rgraphviz)
nodeInfo <- getKEGGnodeData(toyGraph)
nodeType <- sapply(nodeInfo, getType)
makeNodeRenderAttrs <- function(g, label=nodes(g),
                                shape="ellipse", fill="#e0e0e0",...) {
  rv <- list(label=label, shape=shape, fill=fill, ...)
  nA <- nodeRenderInfo(g)
  for(i in seq(along=rv)) {
    if (length(rv[[i]]) == 1) {
      rv[[i]] <- rep(rv[[i]], numNodes(g))
    } else {
      if (length(rv[[i]]) != numNodes(g))
        stop("Attribute vector must have as many elements as 'g' has nodes.")
    }
    names(rv[[i]]) <- nodes(g)
    nA[[ names(rv)[[i]] ]] <- rv[[i]]
  }
  nodeRenderInfo(g) <- nA
  return(g)
}
toyDrawn <- plotKEGGgraph(toyGraph)
toyDrawnRefine <- makeNodeRenderAttrs(toyDrawn, fill=ifelse(nodeType=="gene", "lightblue", "orange"),
                                      shape=ifelse(nodeType=="gene", "ellipse","rectangle"))
```

<a id="fig:01"></a>
```{r vistoyReal, echo=FALSE, warning=FALSE, fig.cap="Visualization of the toy network (incoherent type-1 feed-forward loop, IL1-FFL) created by KGML-ED and parsed by KEGGgraph. The rectangle nodes represent compounds while ellipse ones represent gene/gene products. Solid and dashed red arrows represent activation and expression, whereas blue represent inhibition and repression. Note that self-inhibition of GalS is presented by a node and an edge pointing to its self."}
renderGraph(toyDrawnRefine)
```

Similarly, users could create new or modify existing pathways by modifying KGML
files with tools including KGML-ED, and these pathways can be consequently
parsed, visualized and analyzed consequently by `r Biocpkg("KEGGgraph")` and
other tools for graphs implemented in **Bioconductor**. In this sense
`r Biocpkg("KEGGgraph")` is not only able to study existing biological pathways
constructed by KEGG, but also a tool for modelling pathways and biological
networks in **R**.

# Microarray analysis

KEGG PATHWAY contains pathway maps for nearly 1000 organisms and has been
intensively expanded with pathways including signaling transduction, cellular
processes and human diseases in recent years. As the graph-theory based approach
to this valuable knowledge-base, `r Biocpkg("KEGGgraph")` can be built into
analysis pipelines where the pathway information could be useful, especially
with the graph attributes. To demonstrate this we show one example of microarray
analysis, cooperating with another package `r Biocpkg("SPIA")` [@Tarca2009], a
package implementing the Signaling Pathway Impact Analysis (SPIA) algorithm.

```{r microarrayNote, echo=FALSE, results='asis'}
if(!hasMicroarrayDeps) {
  missingPkgs <- microarrayPkgs[!vapply(microarrayPkgs, requireNamespace,
                                        logical(1), quietly = TRUE)]
  cat(sprintf(paste("**Note**: this section is not evaluated because the",
                    "following suggested package(s) are not installed: %s.",
                    "Install them to reproduce the analysis below.\n"),
              paste(missingPkgs, collapse=", ")))
}
```

The microarray data is decribed in the vignette of `r Biocpkg("SPIA")`, here it
is only briefly summarized: Affemetrix GeneChip (GEO accession: GSE4107),
containing 10 normal samples and 12 colorectal cancer samples, is normalized
with `r Biocpkg("affy")` and `r Biocpkg("limma")` package. The result of
`topTable` in `r Biocpkg("limma")` is used as the input data.

```{r spiaLoad, echo=TRUE, eval=hasMicroarrayDeps, message=FALSE}
if(require(SPIA)) {
  data(colorectalcancer,package="SPIA")
} else {
  load(system.file("extdata/colorectalcancerSPIA.RData", package="KEGGgraph"))
}
head(top)
```

We only consider the probes annotated with EntrezGeneID and differentially
expressed with FDR $p$-value less than 0.05.

```{r spiaTrans, echo=TRUE, eval=hasMicroarrayDeps, message=FALSE}
library(hgu133plus2.db)
x <- hgu133plus2ENTREZID
top$ENTREZ <- unlist(as.list(x[top$ID]))
top <- top[!is.na(top$ENTREZ),]
top <- top[!duplicated(top$ENTREZ),]
tg1 <- top[top$adj.P.Val < 0.05,]
DE_Colorectal <- tg1$logFC
names(DE_Colorectal) <- as.vector(tg1$ENTREZ)
ALL_Colorectal <- top$ENTREZ
```

The *SPIA* algorithm takes as input two vectors: log2 fold changes of
differentially expressed genes (`DE_Colorectal`) and those of all EntrezGeneID
annotated genes (`ALL_Colorectal`), and returns a table of pathways ranked from
the most to the least significant. The results show that Colorectal cancer
pathway (ID:05210) is significantly activated (ranked the 5th of the result
table). To visualize the results, we visualize the pathway with differentially
expressed genes marked with color.

For the first time we could use the `retrieveKGML` function to retrieve the KGML
remotely from the KEGG FTP site.

```{r remoteRetreval, echo=TRUE, eval=TRUE}
tmp <- tempfile()
retrieveKGML(pathwayid="05210", organism="hsa", destfile=tmp)
```

We have attached this file in the `r Biocpkg("KEGGgraph")`, we parse it into
graph and observe that around $30\%$ percent of the genes in the pathway are
differentially expressed. The pathway map (figure) can be found at
https://www.genome.jp/dbget-bin/show_pathway?hsa05210.

```{r spiaPer, echo=TRUE, eval=hasMicroarrayDeps}
colFile <- system.file("extdata/hsa05210.xml",
                       package="KEGGgraph")
g <- parseKGML2Graph(colFile)
deKID <- translateGeneID2KEGGID(names(DE_Colorectal))
allKID <- translateGeneID2KEGGID(ALL_Colorectal)
isDiffExp <- nodes(g) %in% deKID
sprintf("%2.2f%% genes differentially-expressed", mean(isDiffExp)*100)
```

We visualize the pathway with pseudo-colors representing the log2 fold change of
the differentially expressed genes in figure [2](#fig:02).

<a id="fig:02"></a>
```{r spiaVis, echo=FALSE, eval=hasMicroarrayDeps, message=FALSE, warning=FALSE, fig.cap="Overview of colorectal cancer pathway with pseudo-colored nodes representing expression change. Red nodes represented up-regulated mRNAs in cancer samples and blue ones are down-regulated. Grey nodes are detected not to be differentially expressed, yellow ones are not reported (possibly due to under detection threshold). The gradual colors indicate the log2 fold change of the expression - the darker the colors, the larger fold change. One distinct feature is the existence of many non-connected nodes, which could has been omitted when the graph information is discarded and only the membership of the gene is considered."}
library(RColorBrewer)
library(org.Hs.eg.db)
library(RBGL)
library(grid)
ar <- 20
cols <- rev(colorRampPalette(brewer.pal(6, "RdBu"))(ar))
logfcs <- DE_Colorectal[match(nodes(g), deKID)]
names(logfcs) <- nodes(g)
logfcs[is.na(logfcs)] <- 0
incol <- round((logfcs+2)*5); incol[incol>ar] <- ar
undetected <- !nodes(g) %in% allKID
logcol <- cols[incol]; logcol[logfcs==0] <- "darkgrey"; logcol[undetected] <- "yellow"
names(logcol) <- names(logfcs)
nA <- makeNodeAttrs(g, fillcolor=logcol, label="", width=10, height=1.2)
par(mar=c(3,5,0,5), mgp=c(0,0,0))
layout(mat=matrix(c(rep(1,8),2), ncol=1, byrow=TRUE))
plot(g, "dot", nodeAttrs=nA)
image(as.matrix(seq(1,ar)), col=cols, yaxt="n", xaxt="n")
mtext("down-regulation", side=1,  at=0, line=1)
mtext("up-regulation", side=1,  at=1, line=1)
```

```{r spiaSingleNode, echo=FALSE, eval=hasMicroarrayDeps, results='hide', message=FALSE}
gDeg <- degree(g)
gIsSingle <- gDeg[[1]] + gDeg[[2]] == 0
options(digits=3)
gGeneID <- translateKEGGID2GeneID(nodes(g))
gSymbol <-  sapply(gGeneID, function(x) mget(x, org.Hs.egSYMBOL, ifnotfound=NA)[[1]])
isUp <- logfcs > 0
isDown <- logfcs < 0
singleUp <- isUp & gIsSingle
singleDown <- isDown & gIsSingle
```

In the figure [2](#fig:02), we observe that:

* Not all the genes are connected with each other,
  `r val(round(mean(gIsSingle)*100, 1))`%
  (`r val(sum(gIsSingle))`/`r val(length(gIsSingle))`) nodes in the pathway have
  a degree of 0 -- they are not conencted to any other node. There are several
  reasons for this. One of them is the genes with genetic alterations are
  indicated in the pathway map, but not interweaved into the pathway (cf.
  https://www.genome.jp/dbget-bin/show_pathway?hsa05210). Another important
  factor is that in some pathways, especially human disease pathways, KEGG
  records genes involved in the disease but does not provide any edge in that
  pathway, although one can find their interaction partners in other pathways
  (e.g., Grb2 has no edges in colorectal cancer pathway, however in other maps
  including MAPK signaling pathway, both up- and down-stream interactors are
  found). One solution to the later problem is to *merge* (*union*) the related
  pathways together, to this end `r Biocpkg("KEGGgraph")` provides the function
  `mergeKEGGgraphs`.
* From the visualization in figure [2](#fig:02), it is difficult to recognize
  any patterns. To examine the pathway in more details, it is necessary to use
  *Divide and Conquer* strategy, namely to *subset* the graph into subgraphs. To
  this end `r Biocpkg("KEGGgraph")` provides the function `subKEGGgraph`, which
  maintains the KEGG information while subsetting the graph.

Out of `r val(sum(gIsSingle))` nonconnected genes, `r val(sum(singleUp))` is
up-regulated and `r val(sum(singleDown))` downregulated. Notably, out of 10
*Frizzled* homologues (Wnt receptors), four are up-regulated (*FZD2, FZD4, FZD7
and FZD10*), and only one is down-regulated (*FZD5*), in accordance with the
former report [@Vincan2007].

Next we concentrate on upregulated genes, the result given by
`r Biocpkg("SPIA")` package indicates that colorectal pathway is activated.

<a id="fig:03"></a>
```{r spiaSub, echo=FALSE, eval=hasMicroarrayDeps, message=FALSE, warning=FALSE, fig.cap="Up-regulated genes (seven, in red) and their neighborhood genes in colorectal cancer pathway, the color follows the legend in figure 2. Un-reported genes are hidden from the figure."}
ups <- nodes(g)[logfcs > 0]
upNs <- unique(unlist(neighborhood(g, ups, return.self=TRUE)))
upSub <- subKEGGgraph(upNs, g)
upNeighbor <- nodes(upSub)[sapply(neighborhood(upSub, nodes(upSub)), length)>0]
upNeighbor <- setdiff(upNeighbor, nodes(g)[undetected])
upSub <- subKEGGgraph(upNeighbor, upSub)
upSubGID <- translateKEGGID2GeneID(nodes(upSub))
upSymbol <- gSymbol[upSubGID]
upnA <- makeNodeAttrs(upSub, fillcolor=logcol[nodes(upSub)], label=upSymbol, fixedsize=TRUE, width=10, height=10, font=20)
plot(upSub, "dot", nodeAttrs=upnA)
```

We can use degree centrality (the sum of in- and out-degree of each node) as a
measure of the relative importance of the node compared to others. In this
subgraph, 8 nodes have an degree equal or larger than three (in decreasing
order: AKT3, TCF7L1, CCND1, FOS, MAPK3, MAPK1, JUN, MAPK10), and 5 of them are
upreguated to different extent. This reinforces the conclusion of
`r Biocpkg("SPIA")` that the pathway is activated - it seems that the relative
important nodes are up-regulated. Especially we note the upregulation of Jun and
Fos, two transcription factors with many targeting genes, although their
up-stream interactors (MAPK3 and MAPK1) are either not significantly
differentially expressed or slightly down-regulated (MAPK1, logFC=-0.381).

Further analysis could done, for example, by merging the colorectal pathway with
linked pathways (Wnt signaling, apoptosis, etc) and investigate the graph
characteristics of differnetially expressed genes and their links.

# Session info

```{r sessionInfo, echo=FALSE, results='verbatim'}
sessionInfo()
```
